---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# proDD

Differential Detection for Label-free (LFQ) Mass Spectometry Data

The tool fits a probabilistic ropout model to an intensity matrix from from label-free quantification (LFQ). Dropouts in 
LFQ data occur if the protein has a low intensity. Our model takes the non-random missingness into account, by constructing 
a Bayesian hierarchical model. After fitting the model you can sample from the posterior distribution of the means from each
protein and condition. The posterior are a useful element to calculate all kind of statistics/metrics including the probability
that the intensity of a protein in one condition is smaller than in the control (similar to the one-sided p-value).

# Installation

Install the latest version directly from GitHub (make sure that `devtools` is installed)

```{r eval=FALSE, include=TRUE}
devtools::install("const-ae/proDD")
```

# Disclaimer

I am still actively working on the project and although the algorithm is working fine at this point, the API  
might still be subject to change.




