---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# proDD

Differential Detection for Label-free (LFQ) Mass Spectometry Data

The tool fits a probabilistic dropout model to an intensity matrix from from label-free quantification (LFQ). Dropouts in 
LFQ data occur if the protein has a low intensity. Our model takes the non-random missingness into account, by constructing 
a Bayesian hierarchical model. After fitting the model you can sample from the posterior distribution of the means from each
protein and condition. The posterior are a useful element to calculate all kind of statistics/metrics including the probability
that the intensity of a protein in one condition is smaller than in the control (similar to the one-sided p-value).

# Installation

Install the latest version directly from GitHub (make sure that `devtools` is installed)

```{r eval=FALSE, include=TRUE}
devtools::install_github("const-ae/proDD")
```

# Disclaimer

I am still actively working on the project and although the algorithm is working fine at this point, the API  
might still be subject to change.



# Walkthrough

In the following section I will explain how to use the `proDD` package to identify
differential detected protein in label-free mass spectrometry data. I will highlight
all the important functions the package provides.

At first we have to load the `proDD` package and some additional packages
that we will use to plot our data.

```{r}
# My package
library(proDD)

# Packages for plotting
library(ggplot2)
library(pheatmap)
library(viridisLite)
set.seed(1)
```


Next we will load some data. To make our life easier we will use
synthetic data, where we know which proteins have changed and which have not.
For this we will use the `generate_synthetic_data()` function.

```{r}

# The experimental_design is a vector that assignes each sample to one condition
experimental_design <- c("A", "A", "A", "B", "B", "B")

# The generate_synthetic_data can be customized a lot, but here we will 
# use it in its most basic form
syn_data <- generate_synthetic_data(n_rows=1234, experimental_design=experimental_design)

# The data matrix, where non-observed values are coded as NA
X <- syn_data$X

# The columns are the samples and each row is a protein
head(X)
```


To get a better impression of the raw data
we will make a heatmap plot (using the `pheatmap` package). Unfortunately
the `hclust` method that is internally used does not support missing values,
so we will for this plot just replace all missing values with a zero



```{r}
X_for_plotting <- X
X_for_plotting[is.na(X_for_plotting)] <- 0
pheatmap(X_for_plotting,
         main=paste0(round(sum(is.na(X))/prod(dim(X)) * 100), "% missing values"))
```


One important observation is that the missing values do not occur randomly,
but predominantly at low intensities. This can be most clearly be seen, when 
looking at proteins which have some observed and some missing values

```{r}

hist_tmp_data <- data.frame(intensity=c(X), 
           row_has_missing=c(t(apply(X, 1, function(x) rep(any(is.na(x)), ncol(X))))))

ggplot(hist_tmp_data, aes(x=intensity, fill=row_has_missing)) +
    geom_histogram() +
    xlim(12, 32)

```

We conclude from this that there is a certain dropout probability associated
with each latent intensity. At low intensities (e.g. `<15`) it is almost certain
that the values dropout, whereas for high intensities (e.g. `>25`) almost 
no value is missing. We capture this idea using a sigmoidal shaped dropout
curve that looks roughly like this:

```{r}
dropout_curve <- data.frame(intensity=seq(12, 32, length.out=1001))
dropout_curve$dropout_probability <- invprobit(dropout_curve$intensity, rho=18, zeta=-2.5)

ggplot(hist_tmp_data, aes(x=intensity)) +
    geom_histogram(aes(fill=row_has_missing)) +
    geom_line(data=dropout_curve, aes(y=dropout_probability * 600), color="red") +
    xlim(12, 32)

```


Our probabilistic dropout algorithm has two major steps. In the first step
we infer important hyper-parameters of the model using an EM algorithm. The
hyper-parameters that we identify are

* the location and scale of the dropout curve for each sample (called `rho` and `zeta`)
* the overal location of the values (`mu0` and `sigma20`)
* a prior for the protein variances (`nu` and `eta`).

```{r}
# To see the progress while fitting set verbose=TRUE
params <- fit_parameters(X, experimental_design)
params
```

As we can see the method has successfully converged so we can continue. If 
it would not have converged increase `max_iter`. Also we are currently working
on a moderately sized data set, usually a thousand proteins are enough to 
make good estimates of the hyper-parameters, if your dataset has many proteins
you can easily speed up the inference by setting for example `n_subsample=1000`.


Knowing the general distribution of our data, we might be interested how the 
samples are related. Naively we would just calculate the distance matrix
using `dist(t(X))`. But `dist` simply scales up vectors containing missing values,
which is equivalent to some kind of mean imputation, which does not really
make sense as we have seen when we looked where missing values actually occur.

```{r}
naive_dist <- dist(t(X))
pheatmap(naive_dist, cluster_rows=FALSE, cluster_cols = FALSE,
         color=viridisLite::plasma(n=100, direction=-1),
         breaks = seq(30, 60, length.out=101))
```



Instead our package provides a function called `dist_approx` that estimates the
distances and properly takes into account the missing values. But due to the missing
values we cannot be certain of the exact distance. Thus the function returns 
in addition to the best guess of the distance an estimate how variable that 
guess is. 

If we were to call `dist_approx` directly with the `params` object this could bias
our result, because the missing values would be inferred using the group structure. 
But usually this is exactly what we want to check in the first place. Instead we transform the 
parameter object and act as if every sample belonged to the same condition.


```{r}
# Removing condtion information to get unbiased results
dist_proDD2 <- dist_approx(X, transform_parameters(params, rep(1, times=ncol(X))))
# The mean of the sample distance estimates
pheatmap(dist_proDD2$mean, cluster_rows=FALSE, cluster_cols=FALSE,
         color=viridisLite::plasma(n=100, direction=-1),
         breaks = seq(30, 60, length.out=101))

# The standard error of the distance estimates
# In this case the uncertainty is for all distances very small
pheatmap(sqrt(dist_proDD2$var), cluster_rows=FALSE, cluster_cols=FALSE)
```


After making sure that there are no extreme outliers in our data and the heatmap
shows the group structure that we expected, we will continue to infer
the posterior distribution of the mean for each protein and condition.

Those posterior distribution form the basis of the subsequent steps for identifying
differentially detected proteins.


```{r}
# Internally this function uses Stan to sample the posteriors.
# Stan provides a lot of output which you can see by setting verbose=TRUE
posteriors <- sample_protein_means(X, params, verbose=FALSE)
```


Now that we have a good idea what is the latent intensity of each protein
we can go on to identify the differentially detected proteins

```{r}
result <- test_diff(posteriors$A,  posteriors$B)

# The resulting data.frame
head(result)

# The most significant changes
head(result[order(result$pval), ])
```



A popular way to look at such data is to make a volcano plot. Here we will
use the fact that we generated the data to highlight proteins that
were actually changed

```{r}
result$changed <- syn_data$changed

ggplot(result, aes(x=diff, y=-log10(pval), color=changed)) +
    geom_point()
```

We know that 10% of the data was changed and in the volcano plot we can see that
our method does a good job of identifying many of them

An interesting experiment is to look for both conditions how many values were
actually observed and separate the data by this information

```{r}
result$nobs_a <- rowSums(! is.na(X[, experimental_design == "A"]))
result$nobs_b <- rowSums(! is.na(X[, experimental_design == "B"]))

ggplot(result, aes(x=diff, y=-log10(pval), color=changed)) +
    geom_point() +
    facet_grid(nobs_a ~ nobs_b)
```




Using this data we can also make an MA plot, where we color the points
by the number of observations

```{r}
result$label <- paste0(pmax(result$nobs_a, result$nobs_b), "-", pmin(result$nobs_a, result$nobs_b))

ggplot(result, aes(x=mean, y=diff, color=label, shape=changed)) +
    geom_point(size=2)
```






